### 2010-09-21 doc/guide/language.tex
- updated language section
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3394 3e02f596-3131-0410-8352-e04eeaec37cc

The tool \gringo~\cite{gescth07a} is a grounder capable of translating
logic programs provided by users into equivalent ground programs.
The output of \gringo\ can be piped into solver \clasp~\cite{gekanesc07b},
which then computes answer sets.
System \clingo\ internally couples \gringo\ and \clasp, thus,
it takes care of both grounding and solving.
In contrast to \gringo\ outputting ground programs, 
\clingo\ returns answer sets.
Usually logic programs are specified in one or more text files whose names are passed via the command line
in an invocation of either \gringo\ or \clingo.
We below provide a description of constructs belonging to the input language of \gringo\ and \clingo.

@@

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the same form
as a function symbol or constant.
Any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$ where the connective \code{not} corresponds to default negation.
The set of literals $\{L_1,\dots,L_n\}$ is called the body of the rule.
Facts have an empty body.
Throughout this section we will further extend the predicates that can be used 
in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
Furthermore, \gringo{} expects rules to be save, i.e., 
all variables that appear in a rule have to appear in some some positive literal (a literal not preceded by \code{not}) in the body.
If a variable appears positively in some predicate, then we say that this predicate binds the variable.

Intuitively, the head of a rule has to be true whenever all its body literals are true.
In ASP every predicate needs some derivation, i.e., 
a predicate cannot be true if there is no body deriving it.
This implies that only predicates appearing in some head can appear in answer sets.
Furthermore, derivations
\footnote{There are extensions like disjunctions 
that go beyond simple derivability and also require minimality w.r.t. a reduct. 
We will not cover the semantics of such constraints in this guide.}%
have to be acyclic,
a feature that is important to model reachability.
As a simple example consider the program \code{a :- b. b :- a.}
The only answer set to this program is the empty set.
Adding either \code{a.} or \code{b.} to the program results in the answer set $\{a,b\}$.
Finally, note that default negation is ignored when checking for acyclic derivations 
(we do not need a reason for a predicate being false).
Default negation can be used to express choices, e.g., the program \code{a :- not b. b :- not a.} 
has the two answer sets $\{a\}$ and $\{b\}$.
But in practice it is never needed to express choices this way.
For example in the introductory example in Section~\ref{sec:quickstart} we used a cardinality 
constraint, which provides a much more readable way to introduce choices.

A fact has an empty body and thus its associated head predicate is always true 
and will appear in all answer sets.
On the other hand, integrity constraints eliminate answer set candidates.
They are merely tests that discard unwanted answer sets.
That is, there will be no answer sets that satisfy all literals an integrity constraint.
Elaborate examples on the usage of facts, rules, and integrity constraints
will be provided in Section~\ref{sec:examples}.

@@

Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
A disjunction is true if at least one of its atoms is true.
Additionally, logic programs have to satisfy a minimality criterion, 
which we will not detail in this guide.
The simple program \code{a | b.} has the two answer sets $\{a\}$ and $\{b\}$ but
does not admit the answer set ${a,b}$ because it is no minimal model.

In general, the use of disjunction however increases
computational complexity~\cite{eitgot95a}.
This is why \clingo%
\footnote{Run as a monolithic system performing both grounding and solving.}
and solvers like 
\assat~\cite{linzha04a},
\clasp~\cite{gekanesc07b},
\nomorepp~\cite{angelinesc05c},
\smodels~\cite{siniso02a}, and
\smodelscc~\cite{warsch04a}
do not work on disjunctive programs.
Rather,
\claspD~\cite{drgegrkakoossc08a},
\cmodels~\cite{gilima06a,lierler05a}, or
\gnt~\cite{janisesiyo06a}
needs to be used for solving a disjunctive program.%
\footnote{System \dlv~\cite{dlv03a} also deals with disjunctive programs,
  but it uses a different syntax than presented here.}
We thus suggest to use ``choice constructs'' (cf.\ Section~\ref{subsec:gringo:aggregate})
instead of disjunction, unless the latter is required for complexity reasons
(see~\cite{eitpol06a} for an implementation methodology in disjunctive ASP).

@@

\begin{example}\label{ex:arith:fun}
The usage of arithmetic functions is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program,
  obtained after evaluating all arithmetic functions,
  can be inspected by invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/arithf.lp}{arithf.lp}}}
%
\lstinputlisting{examples/arithf.lp}
%
Note that variables~\var{L} and~\var{R} are instantiated to~\const{7} and~\const{2},
respectively, before arithmetic evaluations.
Consecutive and non-separative (e.g., before ``\code{(}'')
spaces can also be dropped,
while spaces after tokens \code{\#div} and \code{\#mod} are mandatory.
Furthermore, the argument of function \code{\#abs}, \code{\#div}, and \code{\#mod} must be enclosed in parentheses.
The four bitwise functions apply to signed integers,
using the complement on two of a negative integer.
\eexample
\end{example}

It is important to note that variables
in the scope of an arithmetic function are not bound by a corresponding atom.
For instance, the rule \code{p(X) :- p(X+1).} is not safe but \code{p(X-1) :- p(X).} is.
Although, the latter might produce an infinite grounding 
and \gringo{} will not necessarily halt when given such an input.

@@

\begin{example}\label{ex:arith:pred}
The usage of comparison predicates is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/arithf.lp}{arithf.lp}}}
%
\lstinputlisting{examples/arithf.lp}
%
The last two lines hint at the fact that arithmetic functions are evaluated
before comparison predicates, so that the latter actually compare integers.

All comparison predicates can also be used with arbitrary ground terms,
as in the next program:%
\marginlabel{%
  As above, invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/symbc.lp}{symbc.lp}}
  yields the unique answer set of the program in terms of facts.}
%
\lstinputlisting{examples/symbc.lp}
%
Integers are compared in the usual way and constants are ordered lexicographically.
Function symbols are compared first using their arity.
If the arity differs, then the name of the function symbol is compared lexicographically.
If again the name differs, then arguments are compared component wise.
Finally, integers are always smaller than constants and constants are always smaller than function symbols.
\eexample
\end{example}

@@

\begin{example}\label{ex:assign}
The next program demonstrates how terms can be assigned to variables:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/assign.lp}{assign.lp}}}%
%
\lstinputlisting{examples/assign.lp}
%
Line~3 contains four assignments, where the right-hand sides directly or indirectly
depend on~\var{X} and~\var{Y}.
These two variables are bound in Line~4 via atoms of predicate \pred{num}/$1$.
Also observe the different usage and role of built-in comparison predicate~\code{==}.
\eexample
\end{example}

\begin{example}\label{ex:unify}
The second program demonstrates the usage of \code{:=},
which allows for terms on the left hand side:
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/unify.lp}{unify.lp}}}%
%
\lstinputlisting{examples/unify.lp}
%
Here the term \code{f(a,X,X+1)} is unified with every function symbol provided by \code{sym}/$1$.
Note the usage of \code{X+1} in the term.
\gringo{} will not try to unify any term containing arithmetic 
but in this example \code{X} occurs also directly as second argument of the argument
and can thus be unified with.
The term $X+1$ is merely a test that deferred and checked later.
For example, the fourth line is equivalent to:
\begin{lstlisting}[firstnumber=6]
unifyf(X) :- f(a,X,Y) := F, sym(F), Y = X + 1.
\end{lstlisting}%
\eexample
\end{example}

Finally, note that assignments to some extend can bind variables.
Of course cyclic assignments cannot bind variables.
For example the rule \code{p(X) :- X = Y, Y = X.} is rejected by \gringo{}.
Either \code{X} or \code{Y} has to be provided by some positive predicate in this case.
Additionally, the unification is restricted to ground terms on the right hand side of the assignment, 
that is all variables on the right hand side have to be by some other predicate.

@@

\begin{example}\label{ex:cond}
The following program uses conditions in a rule body and in a rule head:
%
\lstinputlisting{examples/cond.lp}
%
We are particularly interested in the rules in Line~5 and~6,
instantiated as follows:%
%
\begin{lstlisting}[firstnumber=5]
meet :- available(jane), available(john).
on(mon) | on(tues) | on(wed) | on(thurs) | on(fri) :- meet.
\end{lstlisting}
%
The conjunction in Line~5 is obtained by replacing~\var{X} in
\marginlabel{%
  The reader can reproduce these ground rules by invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/cond.lp}{cond.lp}}}
\code{\pred{available}(\var{X})} with all ground terms~$t$ such that
\code{\pred{person}($t$)} holds, namely, $t=\const{jane}$ and $t=\const{john}$.
Furthermore, the condition in the head of the rule in Line~6 turns into
a disjunction over all ground instances of
\code{\pred{on}(\var{X})} where~\var{X} is substituted by some term~$t$
such that
\code{\pred{day}($t$)} holds.
That is, conditions in the body and in the head of a rule
are expanded to different basic language constructs.

Composite conditions can also be constructed via~``\code{:},''
as in the additional rules:
%
\lstinputlisting[firstnumber=7]{examples/twocond.lp}
%
Observe that we may use the same atom, viz., \code{\pred{day}(\var{X})},
both on the left-hand and on the right-hand side of~``\code{:}.''
Furthermore, negative literals like \code{not \pred{weekend}(\var{X})}
can occur on both sides of a condition.
Note that literals on the right-hand side of a condition are connected conjunctively, that is,
all of them must hold for ground instances of an atom in front
of the condition.
Thus, the instantiated rule in Line~8 looks as follows:
%
\begin{lstlisting}[firstnumber=8]
weekdays :- day(mon), day(tues), day(wed), day(thurs), day(fri).
\end{lstlisting}
%
The atoms in the body of this rule follow from facts, so that
the rule can be simplified to a fact \code{\pred{weekdays}.}
(as done by \gringo).
\eexample
\end{example}

There are three important issues about the correct usage of conditions.
First, all predicates of atoms on the right-hand side of a condition
must be either domain predicates\comment{refer to appendix?},
i.e., predicates that can be completely evaluated during grounding,
or built-in (can be used on the left-hand side but only in a rule body),
which is due to the fact that conditions are evaluated during grounding.

Second, any variable occurring within a condition is considered as \emph{local},
that is, a condition cannot be used to bind variables outside itself.
In turn, variables outside conditions are \emph{global}, and each variable
within an atom in front of a condition must occur on the right-hand side or
be global.
Third, global variables take priority over local ones, that is,
they are instantiated first.
As a consequence, a local variable that also occurs globally is substituted by a term
before the ground instances of a condition are determined.
Hence, the names of local variables must be chosen with care,
making sure that they do not accidentally match the names of global variables.

@@

Optimization statements extend the basic question of
whether a set of atoms is an answer set to
whether it is an optimal answer set.
To support this reasoning mode, \gringo\ and \clingo\ adopt the
optimization statements of \lparse~\cite{lparseManual},
indicated via keywords \const{\#maximize} and \const{\#minimize}.
As an optimization statement does not admit a body, any (local) variable
in it must also occur in an atom (over a domain or built-in predicate)
on the right-hand side of a condition (cf.\ Section~\ref{subsec:gringo:condition})
within the optimization statement.
In multiset notation (square brackets),
weights may be provided as with \const{\#sum} aggregates. %,
In set notation (curly brackets),
duplicates of literals are removed
as with \const{count} aggregates.
Additionally, priorities can be associated with each literal.
A (ground) optimize statement has the form:
\[\mathsf{opt}\,\code{[}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\]
\[\mathsf{opt}\,\code{\char`\{}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\]
where $\mathsf{opt}$ is either \code{\#maximize} or \code{\#minimize} , $L_i$ are literals 
with associates (integer) weights $w_i$ and (integer) priorities $p_i$.

@@

\begin{example}\label{ex:opt}
To illustrate optimization, we consider a hotel booking situation
where we want to choose one among five available hotels.
The hotels are identified via numbers assigned in descending order of stars.
Of course, the more stars a hotel has the more it costs per night.
As an ancillary information, we know that hotel~\const{4} is located
on a main street, which is why we expect its rooms to be noisy.
This knowledge is specified in Line~1--5 of the following program:
%
\lstinputlisting[xrightmargin=-26pt]{examples/opt.lp}
%
Line~6--8 contribute optimization statements in inverse order of significance,
according to which we want to choose the best hotel to book.
The most significant optimization statement in Line~8 states that
avoiding noise is our main priority.
The secondary optimization criterion in Line~7 consists of
minimizing the cost per star.
Finally, the third optimization statement in Line~6 specifies that we want
to maximize the number of stars among hotels that are otherwise indistinguishable.
The optimization statements in Line~6--8 are instantiated as follows:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/opt.lp}{opt.lp}}}
%
\begin{lstlisting}[firstnumber=6,breakindent=0pt,escapechar=&]
#maximize [ hotel(1)=5@1, hotel(2)=4@1, &\\&            hotel(3)=3@1, hotel(4)=3@1, hotel(5)=2@1 ].
#minimize [ hotel(1)=34@2, hotel(2)=35@2, &\\&            hotel(3)=30@2, hotel(4)=25@2, hotel(5)=30@2 ].
#minimize [ noisy=1@3 ].
\end{lstlisting}
If we now use \clasp\ to compute an optimal answer set,%
\marginlabel{%
  To compute the unique optimal answer set, invoke:\\
  \code{\mbox{~}gringo \attach{examples/opt.lp}{opt.lp} | \textbackslash\\
        \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \attach{examples/opt.lp}{opt.lp}}}
we find that hotel~\const{4} is not eligible because it implies \pred{noisy}.
Thus, hotel~\const{3} and~\const{5} remain as optimal w.r.t.\ the second most
significant optimization statement in Line~7.
This tie is broken via the least significant optimization statement in Line~6
because hotel~\const{3} has one star more than hotel~\const{5}.
We thus decide to book hotel~\const{3} offering~\const{3} stars
to cost~\const{90} per night.
\eexample
\end{example}

@@

\paragraph{Comments.}
To keep records of the contents of a logic program,
a logic program file may include comments.
A comment until the end of a line is initiated by symbol ``\code{\%},''
and a comment within one or over multiple lines is enclosed
by ``\code{\%*}'' and ``\code{*\%}.''
As an abstract example, consider:
%
\begin{lstlisting}[numbers=none,escapechar=@]
@\textit{logic program}@  %* enclosed comment *%  @\textit{logic program}@
@\textit{logic program}@  % comment till end of line
@\textit{logic program}@
%*
comment over multiple lines
*%
@\textit{logic program}@
\end{lstlisting}

@@

In order to selectively include the atoms of a certain predicate in the output,
one may use the \code{\#show} declarative.
Here are some examples:
%
\begin{lstlisting}[numbers=none]
#show p/3. % Include all atoms of predicate p/3 in output
#showp(X,Y) : q(X). % Supress p/3 if the condition holds
\end{lstlisting}
%
A typical usage of \code{\#hide} and \code{\#show} is to hide all predicates
via ``\code{\#hide.}'' and to selectively re-add atoms of certain predicates
\code{p/$n$} to the output via ``\code{\#show p/$n$.}''

\paragraph{Constant Replacement.}
Constants appearing in a logic program may actually be placeholders for
concrete values to be provided by a user.
An example of this will be given in Section~\ref{subsec:ex:color}.
Via the \code{\#const} declarative,
one may define a default value to be inserted for a constant.
Such a default value can still be overridden via command line option
\code{--const} (cf.\ Section~\ref{subsec:opt:gringo}).
Syntactically, \code{\#const} must be followed by an assignment having
a (symbolic) constant on the left-hand side and a term on the right-hand side.
Some exemplary \code{\#const} declarations are:
%
\begin{lstlisting}[numbers=none]
#const x = 42.
#const y = f(x,h).
\end{lstlisting}%
Note that (for efficiency reasons) constant declarations are order dependent.
In the example above \code{x} would be replaced by \code{42}
but when reversing the directives this would no longer be the case.

\paragraph{Domain Declarations.}
Usually, variable names are local to a rule,
where they must be bound via appropriate atoms.
This locality can be undermined by using
\code{\#domain} declarations 
that globally associate variable names to atoms.
An associated atom is then simply added to the body of a rule in which
such a predefined variable name occurs in.
The following is a made-up example:
%
\begin{lstlisting}[numbers=none]
p(1,1). p(1,2).
#domain p(X,Y).
#domain p(Y,Z).
q(Z,X) :- not p(Z,X).
\end{lstlisting}
%
The above program is a priori not safe because
variables~\var{X} and~\var{Z} are unbound in the last rule.
However, as they belong to \code{\#domain} declarations,
\gringo\ and \clingo\ expand the last rule to:
%
\begin{lstlisting}[numbers=none]
q(Z,X) :- p(X,Y), p(Y,Z), not p(Z,X).
\end{lstlisting}
%
Observe that the resulting program is safe.
Nevertheless, we suggest not to use domain statements because 
in ASP it is common to use very short variable names and
using domain statements will likely result in name clashes
and undesired behavior.

>>>>

The tool \gringo~\cite{gescth07a} is a grounder capable of translating
logic programs provided by users into equivalent ground programs.
The output of \gringo\ can be piped into solver \clasp~\cite{gekanesc07b},
which then computes answer sets.
System \clingo\ internally couples \gringo\ and \clasp, thus,
it takes care of both grounding and solving.
In contrast to \gringo\ outputting ground programs, 
\clingo\ returns answer sets.

Usually logic programs are specified in one or more text files whose names are passed via the command line
in an invocation of either \gringo\ or \clingo.
We below provide a description of constructs belonging to the input language of \gringo\ and \clingo.

@@

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the same form
as a function symbol or constant.
Any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$ where the connective \code{not} corresponds to default negation.
The set of literals $\{L_1,\dots,L_n\}$ is called the body of the rule.
Facts have an empty body.
Throughout this section we further extend the predicates that can be used 
in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
Furthermore, \gringo{} expects rules to be safe, i.e., 
all variables that appear in a rule have to appear in some positive literal (a literal not preceded by \code{not}) in the body.
If a variable appears positively in some predicate, then we say that this predicate binds the variable.

Intuitively, the head of a rule has to be true whenever all its body literals are true.
In ASP every atom needs some derivation, i.e., 
an atom cannot be true if there is no rule deriving it.
This implies that only atoms appearing in some head can appear in answer sets.
Furthermore, derivations%
\footnote{There are extensions like disjunctions 
that go beyond simple derivability and also require minimality w.r.t. a reduct. 
We do not cover the semantics of such constraints in this guide.}%
have to be acyclic,
a feature that is important to model reachability.
As a simple example, consider the program \code{a :- b. b :- a.}
The only answer set to this program is the empty set.
Adding either \code{a.} or \code{b.} to the program results in the answer set $\{a,b\}$.
Finally, note that default negation is ignored when checking for acyclic derivations 
(we do not need a reason for an atom being false).
Default negation can be used to express choices, e.g., the program \code{a :- not b. b :- not a.} 
has the two answer sets $\{a\}$ and $\{b\}$.
But in practice it is never needed to express choices this way.
For example in the introductory example in Section~\ref{sec:quickstart} we used a cardinality 
constraint, which provides a much more readable way to introduce choices.
\com{This paragraph contains a lot of information. Could someone help me making it more readable?}

A fact has an empty body and thus its associated head predicate is always true 
and appears in all answer sets.
On the other hand, integrity constraints eliminate answer set candidates.
They are merely tests that discard unwanted answer sets.
That is, there are no answer sets that satisfy all literals an integrity constraint.
Elaborate examples on the usage of facts, rules, and integrity constraints
are provided in Section~\ref{sec:examples}.

@@

Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
A disjunction is true if at least one of its atoms is true.
Additionally, logic programs have to satisfy a minimality criterion, 
which we do not detail in this guide.
The simple program \code{a | b.} has the two answer sets $\{a\}$ and $\{b\}$ but
does not admit the answer set ${a,b}$ because it is no minimal model.

In general, the use of disjunction however increases
computational complexity~\cite{eitgot95a}.
This is why \clingo%
\footnote{Run as a monolithic system performing both grounding and solving.}
and solvers like 
\assat~\cite{linzha04a},
\clasp~\cite{gekanesc07b},
\nomorepp~\cite{angelinesc05c},
\smodels~\cite{siniso02a}, and
\smodelscc~\cite{warsch04a}
do not work on disjunctive programs.
Rather,
\claspD~\cite{drgegrkakoossc08a},
\cmodels~\cite{gilima06a,lierler05a}, or
\gnt~\cite{janisesiyo06a}
need to be used for solving a disjunctive program.%
\footnote{System \dlv~\cite{dlv03a} also deals with disjunctive programs,
  but it uses a different syntax than presented here.}
We thus suggest to use ``choice constructs'' (cf.\ Section~\ref{subsec:gringo:aggregate})
instead of disjunction, unless the latter is required for complexity reasons
(see~\cite{eitpol06a} for an implementation methodology in disjunctive ASP).

@@

\begin{example}\label{ex:arith:fun}
The usage of arithmetic functions is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program,
  obtained after evaluating all arithmetic functions,
  can be inspected by invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/arithf.lp}{arithf.lp}}}
%
\lstinputlisting{examples/arithf.lp}
%
Note that variables~\var{L} and~\var{R} are instantiated to~\const{7} and~\const{2},
respectively, before arithmetic evaluations.
Consecutive and non-separative (e.g., before ``\code{(}'')
spaces can also be dropped,
while spaces after tokens \code{\#div} and \code{\#mod} are mandatory.
Furthermore, the argument of function \code{\#abs}, \code{\#div}, and \code{\#mod} must be enclosed in parentheses.
The four bitwise functions apply to signed integers,
using the two's complement of a negative integer.
\eexample
\end{example}

It is important to note that variables
in the scope of an arithmetic function are not bound by a corresponding atom.
For instance, the rule \code{p(X) :- p(X+1).} is not safe but \code{p(X-1) :- p(X).} is.
Although, the latter might produce an infinite grounding 
and \gringo{} not necessarily halts when given such an input.

@@

\begin{example}\label{ex:arith:pred}
The usage of comparison predicates is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/arithc.lp}{arithc.lp}}}
%
\lstinputlisting{examples/arithc.lp}
%
The last two lines hint at the fact that arithmetic functions are evaluated
before comparison predicates, so that the latter actually compare integers.

All comparison predicates can also be used with arbitrary ground terms,
as in the next program:%
\marginlabel{%
  As above, invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/symbc.lp}{symbc.lp}}
  yields the unique answer set of the program in terms of facts.}
%
\lstinputlisting{examples/symbc.lp}
%
Integers are compared in the usual way and constants are ordered lexicographically.
Function symbols are compared first using their arity.
If the arity differs, then the name of the function symbol is compared lexicographically.
If again the name differs, then arguments are compared component wise.
Finally, integers are always smaller than constants and constants are always smaller than function symbols.
\eexample
\comment{add \#sup and \#inf?}
\end{example}

@@

\begin{example}\label{ex:assign}
The next program demonstrates how terms can be assigned to variables:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/assign.lp}{assign.lp}}}%
%
\lstinputlisting{examples/assign.lp}
%
Line~3 contains four assignments, where the right-hand sides directly or indirectly
depend on~\var{X} and~\var{Y}.
These two variables are bound in Line~5 via atoms of predicate \pred{num}/$1$.
Also observe the different usage and role of built-in comparison predicate~\code{==}.
\eexample
\end{example}

\begin{example}\label{ex:unify}
The second program demonstrates the usage of \code{:=},
which allows for terms on the left hand side:
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/unify.lp}{unify.lp}}}%
%
\lstinputlisting{examples/unify.lp}
%
Here the term \code{f(a,X,X+1)} is unified with every function symbol provided by \code{sym}/$1$.
Note the usage of \code{X+1} in the term.
\gringo{} does not try to unify any term containing arithmetic 
but in this example \code{X} occurs also directly as second argument of the argument
and can thus be unified with.
The term $X+1$ is merely a test that is deferred and checked later.
For example, the fourth line is equivalent to:
\begin{lstlisting}[firstnumber=6]
unifyf(X) :- f(a,X,Y) := F, sym(F), Y == X + 1.
\end{lstlisting}%
\eexample
\end{example}

Finally, note that assignments to some extent can bind variables.
Of course cyclic assignments cannot bind variables.
For example the rule \code{p(X) :- X = Y, Y = X.} is rejected by \gringo{}.
Either \code{X} or \code{Y} has to be provided by some positive predicate in this case.
Additionally, unification is restricted to ground terms on the right hand side of the assignment, 
that is, all variables on the right hand side have to be bound by some other predicate.

@@

\begin{example}\label{ex:cond}
The following program uses conditions in a rule body and in a rule head:
%
\lstinputlisting{examples/cond.lp}
%
We are particularly interested in the rules in Line~5 and~6,
instantiated as follows:%
%
\begin{lstlisting}[firstnumber=5]
meet :- available(jane), available(john).
on(mon) | on(tue) | on(wed) | on(thu) | on(fri) :- meet.
\end{lstlisting}
%
The conjunction in Line~5 is obtained by replacing~\var{X} in
\marginlabel{%
  The reader can reproduce these ground rules by invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/cond.lp}{cond.lp}}}
\code{\pred{available}(\var{X})} with all ground terms~$t$ such that
\code{\pred{person}($t$)} holds, namely, $t=\const{jane}$ and $t=\const{john}$.
Furthermore, the condition in the head of the rule in Line~6 turns into
a disjunction over all ground instances of
\code{\pred{on}(\var{X})} where~\var{X} is substituted by some term~$t$
such that
\code{\pred{day}($t$)} holds.
That is, conditions in the body and in the head of a rule
are expanded to different basic language constructs.

Composite conditions can also be constructed via~``\code{:},''
as in the additional rules:
%
\lstinputlisting[firstnumber=7]{examples/twocond.lp}
%
Observe that we may use the same atom, viz., \code{\pred{day}(\var{X})},
both on the left-hand and on the right-hand side of~``\code{:}.''
Furthermore, negative literals like \code{not \pred{weekend}(\var{X})}
can occur on both sides of a condition.
Note that literals on the right-hand side of a condition are connected conjunctively, that is,
all of them must hold for ground instances of an atom in front
of the condition.
Thus, the instantiated rule in Line~8 looks as follows:
%
\begin{lstlisting}[firstnumber=8]
weekdays :- day(mon), day(tue), day(wed), day(thu), day(fri).
\end{lstlisting}
%
The atoms in the body of this rule follow from facts, so that
the rule can be simplified to a fact \code{\pred{weekdays}.}
(as done by \gringo).
\eexample
\end{example}

Note that there are three important issues about the correct usage of conditions:
\begin{enumerate}
\item All predicates of atoms on the right-hand side of a condition
must be either domain predicates\comment{refer to appendix?},
i.e., predicates that can be completely evaluated during grounding,
or built-in, which is due to the fact that conditions are evaluated during grounding.
\item Any variable occurring within a condition is considered as \emph{local},
that is, a condition cannot be used to bind variables outside the condition.
In turn, variables outside conditions are \emph{global}, and each variable
within an atom in front of a condition must occur on the right-hand side or
be global.
\item Global variables take priority over local ones, that is,
they are instantiated first.
As a consequence, a local variable that also occurs globally is substituted by a term
before the ground instances of a condition are determined.
Hence, the names of local variables must be chosen with care,
making sure that they do not accidentally match the names of global variables.
\end{enumerate}

@@

Optimization statements extend the basic question of
whether a set of atoms is an answer set to
whether it is an optimal answer set.
To support this reasoning mode, \gringo\ and \clingo\ adopt the
optimization statements of \lparse~\cite{lparseManual},
indicated via keywords \const{\#maximize} and \const{\#minimize}.
As an optimization statement does not admit a body, any (local) variable
in it must also occur in an atom (over a domain or built-in predicate)
on the right-hand side of a condition (cf.\ Section~\ref{subsec:gringo:condition})
within the optimization statement.
In multiset notation (square brackets),
weights may be provided as with \const{\#sum} aggregates. %,
In set notation (curly brackets),
duplicates of literals are removed
as with \const{count} aggregates.
Additionally, priorities can be associated with each literal.
A (ground) optimize statement has the form:
\[\mathsf{opt}\,\code{[}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\] 
\[\mathsf{opt}\,\code{\char`\{}\,L_1@p_1,\dots,L_n@p_n\,\code{\char`\}}\]
where $\mathsf{opt}$ is either \code{\#maximize} or \code{\#minimize} , $L_i$ are literals 
with associates (integer) weights $w_i$ and (integer) priorities $p_i$.

@@

\begin{example}\label{ex:opt}
To illustrate optimization, we consider a hotel booking situation
where we want to choose one among five available hotels.
The hotels are identified via numbers assigned in descending order of stars.
Of course, the more stars a hotel has, the more it costs per night.
As an ancillary information, we know that hotel~\const{4} is located
on a main street, which is why we expect its rooms to be noisy.
This knowledge is specified in Line~1--5 of the following program:
%
\lstinputlisting[xrightmargin=-26pt]{examples/opt.lp}
%
Line~6--8 contribute optimization statements in inverse order of significance,
according to which we want to choose the best hotel to book.
The most significant optimization statement in Line~8 states that
avoiding noise is our main priority.
The secondary optimization criterion in Line~7 consists of
minimizing the cost per star.
Finally, the third optimization statement in Line~6 specifies that we want
to maximize the number of stars among hotels that are otherwise indistinguishable.
The optimization statements in Line~6--8 are instantiated as follows:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/opt.lp}{opt.lp}}}
%
\begin{lstlisting}[firstnumber=6,breakindent=0pt,escapechar=&]
#maximize [ hotel(1)=5@1, hotel(2)=4@1, &\\&            hotel(3)=3@1, hotel(4)=3@1, hotel(5)=2@1 ].
#minimize [ hotel(1)=34@2, hotel(2)=35@2, &\\&            hotel(3)=30@2, hotel(4)=25@2, hotel(5)=30@2 ].
#minimize [ noisy=1@3 ].
\end{lstlisting}
If we now use \clasp\ to compute an optimal answer set,%
\marginlabel{%
  To compute the unique optimal answer set, invoke:\\
  \code{\mbox{~}gringo \attach{examples/opt.lp}{opt.lp} | \textbackslash\\
        \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \attach{examples/opt.lp}{opt.lp}}}
we find that hotel~\const{4} is not eligible because it implies \pred{noisy}.
Thus, hotel~\const{3} and~\const{5} remain as optimal w.r.t.\ the second most
significant optimization statement in Line~7.
This tie is broken via the least significant optimization statement in Line~6
because hotel~\const{3} has one star more than hotel~\const{5}.
We thus decide to book hotel~\const{3} offering~\const{3} stars
to cost~\const{90} per night.
\eexample
\end{example}

@@

\paragraph{Comments.}
To keep records of the contents of a logic program,
a logic program file may include comments.
A comment until the end of a line is initiated by symbol ``\code{\%},''
and a comment within one or over multiple lines is enclosed
in ``\code{\%*}'' and ``\code{*\%}.''
As an abstract example, consider:
%
\begin{lstlisting}[numbers=none,escapechar=@]
@\textit{logic program}@  %* enclosed comment *%  @\textit{logic program}@
@\textit{logic program}@  % comment till end of line
@\textit{logic program}@
%*
comment over multiple lines
*%
@\textit{logic program}@
\end{lstlisting}

@@

In order to selectively include the atoms of a certain predicate in the output,
one may use the \code{\#show} declarative.
Here are some examples:
%
\begin{lstlisting}[numbers=none]
#show p/3. % Include all atoms of predicate p/3 in output
#show(X,Y) : q(X). % Include p/3 if the condition holds
\end{lstlisting}
%
A typical usage of \code{\#hide} and \code{\#show} is to hide all predicates
via ``\code{\#hide.}'' and to selectively re-add atoms of certain predicates
\code{p/$n$} to the output via ``\code{\#show p/$n$.}''

\paragraph{Constant Replacement.}
Constants appearing in a logic program may actually be placeholders for
concrete values to be provided by a user.
An example of this is given in Section~\ref{subsec:ex:color}.
Via the \code{\#const} declarative,
one may define a default value to be inserted for a constant.
Such a default value can still be overridden via command line option
\code{--const} (cf.\ Section~\ref{subsec:opt:gringo}).
Syntactically, \code{\#const} must be followed by an assignment having
a (symbolic) constant on the left-hand side and a term on the right-hand side.
Some exemplary constant declarations are:
%
\begin{lstlisting}[numbers=none]
#const x = 42.
#const y = f(x,h).
\end{lstlisting}%
Note that (for efficiency reasons) constant declarations are order dependent.
In the example above \code{x} would be replaced by \code{42}
but when reversing the directives this would no longer be the case.

\paragraph{Domain Declarations.}
Usually, variable names are local to a rule,
where they must be bound via appropriate atoms.
This locality can be undermined by using
\code{\#domain} declarations 
that globally associate variable names to atoms.
An associated atom is then simply added to the body of a rule in which
such a predefined variable name occurs in.
The following is a made-up example:
%
\begin{lstlisting}[numbers=none]
p(1,1). p(1,2).
#domain p(X,Y).
#domain p(Y,Z).
q(Z,X) :- not p(Z,X).
\end{lstlisting}
%
The above program is a priori not safe because
variables~\var{X} and~\var{Z} are unbound in the last rule.
However, as they belong to \code{\#domain} declarations,
\gringo\ and \clingo\ expand the last rule to:
%
\begin{lstlisting}[numbers=none]
q(Z,X) :- p(X,Y), p(Y,Z), not p(Z,X).
\end{lstlisting}
%
Observe that the resulting program is safe.
Nevertheless, we suggest not to use domain statements because 
in ASP it is common to use very short variable names and
using domain statements likely results in name clashes
and undesired behavior.
